---
title: "project2"
author: "Haoming Chen"
date: "2025-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

I choose three datasets from the discussion forum, which are airbooking_table, film_review_table and students_grades_table.
They are all in a wide format and I am going to tidy and transform them into long format using tidyverse library. 

## 1. Air Booking Table 

```{r air_booking}
# The first table is about air booking agencies.
ab <- read.csv('https://raw.githubusercontent.com/vincent-usny/607_project2/refs/heads/main/airbooking_table.csv')

ab <- ab %>%
  pivot_longer(
    cols = c(Winter, Spring, Summer, Fall),
    names_to = "Season",
    values_to = "Price"
  )
```

```{r}
# compare avg_price by season among agencies
ab %>%
  group_by(Agency,Season) %>%
  summarise(avg_price = round(mean(Price))， .groups = "drop") %>%
  arrange(desc(avg_price))
```

```{r}
# compare the services' avg_price among agencies
ab %>%
  group_by(Agency, Service)%>%
  summarise(avg_price = mean(Price), .groups = "drop") %>%
  arrange(desc(avg_price))
```

```{r plot_ab}
ggplot(ab, aes(x = Season, y = Price, color = Service, group = Service)) + geom_line() + facet_wrap(~Agency)
```

# 2. Film Review Table

```{r}
film <- read.csv('https://raw.githubusercontent.com/vincent-usny/607_project2/refs/heads/main/film_review_table.csv')

film <- film %>%
  pivot_longer(
    cols = c(Daniel,Eleana,Susan,Winnie,Aiden),
    names_to = "Reviewer",
    values_to = "Rating"
  )
```

```{r}
# average rating per film
film %>%
  group_by(Film) %>%
  summarise(avg_rating = round(mean(Rating),2), .group = "drop") %>%
  arrange(desc(avg_rating))
```

```{r}
ggplot(film, aes(x = Film, y = Reviewer, fill = Rating)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
# This tile plot shows each reviewer's rating toward each film.
# We can also see some reviewers tend to give higer ratings. 
```

## 3. Students Grades Table

```{r}
grades <- read.csv('https://raw.githubusercontent.com/vincent-usny/607_project2/refs/heads/main/students_grades_table.csv')

grades <- grades %>%
  pivot_longer(
    cols = c(Math, Science, English),
    names_to = "Subject",
    values_to = "Score"
  )
```

```{r}
# compare each student's overall avg_score each year
grades %>%
  group_by(Student, Year) %>%
  summarise(avg_score = round(mean(Score),1), .groups = "drop") %>%
  arrange(desc(avg_score))
```

```{r}
# rank each subject's avg_score among students 
grades %>%
  group_by(Student, Subject) %>%
  summarise(avg_score = round(mean(Score),1), .groups = "drop") %>%
  arrange(Subject,desc(avg_score))
```

```{r plot}
ggplot(grades, aes(x = Year, y = Score, color = Subject, group = Subject)) + 
  geom_line() + 
  facet_wrap(~Student)

# we can see David's overall performance is the worst while Caro's is the best.
# Alice is the second and Bob is the third. It's quite clear.
```

## Conclusion

These three wide datasets are successfully tidied and transformed into long formats.
The data is analyzed in some levels while I think the line chart and the tile 
chart perform pretty well to describe the data.
